\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-12-29 archive.tex
lots of improvements

@@

\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=white!5!white] (-1,1) -- (13,1) -- (13,-4) -- (-1,-4) -- (-1,1);
    \node at (5.5,1.25) {\textsc{\textbf{Wake}}};

@@

  \end{tikzpicture}
  \caption{}\label{threeCycles}
  \end{figure}

    \newcommand{\name}[2]{
      \mcode{{\color{\inventionColor}$f_{#1}$}$($#2$)\,=\,$}
    }
    \newcommand{\noArguments}[1]{
      \mcode{{\color{\inventionColor}$f_{#1}\,=\,$}}
    }
    \newcommand{\use}[1]{{\color{\inventionColor}$f_{#1}$}}
    \renewcommand{\horizontalSpacing}{4.2cm}
    \renewcommand{\verticalSpacing}{2cm}

    \begin{scope}[shift={(0,8)}]
      \task{1}{2,3}{
        \textsc{Take if $\geq 5$}\\
        \mcode{[1 4 5 8]$\to$[5 8]}\\
        \mcode{[9 8 3 8]$\to$[9 8 8]}\\
        \mcode{[3 4 5 6]$\to$[5 6]}\\
        \mcode{[6 5 4 3]$\to$[6 5]}
      }{\footnotesize $f(\text{\mcode{z}}) = $\mcode{(${\color{\inventionColor}f_0}$ ($\lambda$ (x)}\\\hspace{0.5cm}\mcode{ (> x ${\color{\inventionColor}f_4}$)) z)}}
      \nextTask{2}{t1}{
        \textsc{Reverse}\\
        \mcode{[9 2 4]$\to$[4 2 9]}\\
        \mcode{[4 3]$\to$[3 4]}\\
        \mcode{[3 4 5 6]$\to$[6 5 4 3]}\\
        \mcode{[]$\to$[]}
      }{\footnotesize $f(\text{\mcode{z}}) = $\mcode{(fold ($\lambda$ (x a)}\\\hspace{0.25cm}\mcode{ (${\color{\inventionColor}f_3}$ x a)) nil z)}}

      \vinvention{filter}{p1}{
        \name{0}{p$,$z}\mcode{\\(fold ($\lambda$ (x a)\\ (if (p x) (cons x a) a))\\ nil z)} 
      }{higher-order `filter'}

      \vinvention{three}{filter}{\footnotesize \noArguments{1}\\\mcode{(+ 1 (+ 1 1))}}{The number `3'}
      \vinvention{prefix}{three}{\name{2}{n$,$z}\\\mcode{(map ($\lambda$ (i) (index i z)) (range (+ 1 n)))}}{First $n$ elements}
      \hinvention{single}{filter}{\name{3}{z$,$l}\\\mcode{(fold ($\lambda$ (x y)\\(cons x y))\\ z l)}}{Append element}
      \vinvention{for}{single}{\noArguments{4}\\\mcode{(+ 1 ${\color{\inventionColor}f_1}$)}}{The number `4'}
      \vinvention{Max}{for}{\name{5}{l}\\\mcode{%
          (car (${\color{\inventionColor}f_0}$ ($\lambda$ (x) (nil? (${\color{\inventionColor}f_0}$ l ($\lambda$ (z) (> z x))))) l))}}{max of list}
      \hinvention{largest}{single}{\name{6}{l$,$n}\\\mcode{(${\color{\inventionColor}f_5}$ (${\color{\inventionColor}f_0}$ ($\lambda$ (x) \\(> n (length (${\color{\inventionColor}f_0}$ ($\lambda$ (y) (> x y)) l)))) l))}}{$n^{\text{th}}$ largest element}
      \vinvention{count}{largest}{\name{7}{p$,$l}\\\mcode{(length\\ (${\color{\inventionColor}f_0}$ ($\lambda$ (y) (= x y)) l))}}{count}
      \vinvention{weird}{count}{\name{8}{p$,$l}\mcode{
          (map ($\lambda$ (x) (if (p x) (+ 1 x) 0)))}}{Higher-order incrementer}

      \node[rotate=90,align=center] at ([xshift=-2cm]t1.center) {\textbf{\textsc{Tasks}}\\};
      \node[rotate=90,align=center] at ([xshift=-2cm]p1.center) {\texttt{\textbf{Progs.}}\\};
      \node[rotate=90,align=center] at ([xshift=-2cm]three.center) {\textbf{\textsc{Domain Specific Language}}\\
        \small Learned routines in {\color{orange}orange}};

      \draw[->] ([yshift=0.35cm,xshift=-0.3cm]p1.south) to[out=250] (filter.north);
      \draw[->] ([yshift=0.0cm,xshift=0.3cm]p1.south) to[out=350,in=100] (for.west);
      \draw[->] ([xshift=-0.7cm]p2.south) -- (single.north);
      \draw[->] ([xshift=-1.3cm,yshift=0.6cm]p3.south) to[out=240] (largest.north);

    \node[align=center,rotate=90] at ([xshift=-2cm]t1.center) {\textbf{\textsc{Tasks}}\\};
    \node(programLabel)[align=center,rotate=90] at ([xshift=-2cm]p1.center) {\texttt{\textbf{Progs.}}\\};
    \node[rotate=90,align=center] at ([xshift=-2cm]pu.center) {\textbf{\textsc{Domain Specific Language}}\\
      \small Learned routines in {\color{orange}orange}};

    \draw [dashed] ([xshift=0.5cm,yshift=-0.25\verticalSpacing]programLabel.east) -- ([xshift=15cm,yshift=-0.25\verticalSpacing]programLabel.east);
    \node[align=left,anchor=center] at ([xshift=14cm,yshift=-0.3\verticalSpacing]programLabel.east) {\large $\uparrow$\;system inputs\\\\\\\large $\downarrow$\;system outputs};
  \end{tikzpicture}
  \caption{Two of the six domains we apply \system to. Agent observes tasks (top rows) which it solves by writing programs (middle rows) while jointly growing a library (DSL; bottom rows). Learned DSLs rediscover multiple higher-order functions (\code{filter} for list functions and rotational symmetry for generative graphics). Learned DSL components call each other (arrows).}  
\end{figure}

An age-old dream within AI is a machine that learns and reasons by
writing its own programs.  This vision stretches back to the
1960's~\cite{solomonoff1964formal} and cuts to the core of much of
what it would take to build machines that learn and think like humans.
Computational models of cognition often explain the flexibility and
richness of human thinking in terms of program learning: from everyday
thinking and problem solving (motor program induction as an account of
recognition and generation of handwriting and
speech~\cite{lake2015human}; functional programs as a model of natural
language semantics~\cite{SOMETHING}) to learning problems that unfold
over longer developmental time scales: the child's acquisition of
intuitive theories (of kinship, taxonomy, etc.)~\cite{Ullman2012} and
natural language grammar~\cite{DBLP:journals/cogsr/SchmidK11}, to name
just a few.  An outstanding challenge, however, is to engineer
program-learners that display the same level of domain-generality as
the humans they are meant to model.

We contribute a program-induction system that learns the
domain-specific concepts (DSL) while jointly learning how to use those
concepts.  This joint learning problem models two complementary
notions of domain expertise: (1) domain experts have at their disposal
a powerful, yet specialized repertoire of concepts and abstractions
(analogous to the DSL) while also (2) having accurate intuitions about
when and how to use those concepts to solve new problems.

We call our system `DreamCoder' because it acquires these two kinds of
expertise through a novel kind of wake/sleep or `dream'
learning~\cite{hinton1995wake}, iterating through a wake cycle --
where it solves problems by writing programs -- and a pair of sleep
cycles, both of which are loosely biologically inspired by actual
human sleep.  The first sleep cycle, which we refer to as
\textbf{memory consolidation}, grows the DSL by replying experiences
from waking and consolidating them into new code abstractions.  This
cycle is inspired by the formation of abstractions during sleep memory
consolidation~\cite{DUDAI201520}.  The second sleep cycle, which we
refer to as \textbf{dreaming}, improves the agents knowledge of how to
write programs by training a neural network on replayed experiences as
well as `fantasies', or samples, from the DSL.  These two kinds of
dreams are inspired by the distinct episodic replay and hallucination
components of dream sleep~\cite{fosse2003dreaming}.

This dream-learning architecture brings together two lines of prior work.
One line of work
considers the problem of learning new concepts, abstractions, or
`options' from
experience~\cite{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,solomonoff1989system,DBLP:journals/corr/abs-1110-5667,stolle2002learning},
while the other line of work considers the problem of learning how to
deploy those concepts
efficiently~\cite{devlin2017robustfill,balog2016deepcoder,NGDS}.  Our
goal with DreamCoder is to show that the combination of these ideas is
uniquely powerful, and pushes us toward program-writing systems that,
like human learners, autonomously acquire the expertise needed to
navigate a new domain of problems.

\section{Deep Wake/Sleep Program Induction}

\system~takes as its goal to acquire domain-specific expertise, which
it learns by solving a collection of programming \textbf{tasks} --
typically about a hundred.  It alternatingly finds programs that solve
tasks (Wake -- Figure~\ref{threeCycles} top); improving its DSL by
discovering and reusing domain-specific subroutines (Memory
Consolidation -- Figure~\ref{threeCycles} left); and training a neural
network that efficiently guides search for programs in the DSL
(Dreaming -- Figure~\ref{threeCycles} right).  The learned DSL
effectively encodes a prior on programs likely to solve tasks in the
domain, while the neural net looks at the example input-output pairs
for a specific task and produces a ``posterior'' for programs likely
to solve that specific task (Figure~\ref{threeCycles} middle).  The
neural network thus functions as a \textbf{recognition model}
supporting a form of approximate Bayesian program induction, jointly
trained with a \textbf{generative model} for programs encoded in the
DSL, in the spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL (and hence the search space for programs)
expands.

%% \section{Introduction: NIPS}

%% Much of everyday human thinking and learning can be understood in
%% terms of program induction: constructing a procedure that maps inputs
%% to desired outputs, based on observing example input-output pairs.
%% People can induce programs flexibly across many different domains,
%% often from just one or a few examples.  For instance, if shown that a
%% text-editing program should map ``Jane Morris Goodall'' to
%% ``J. M. Goodall'', we can guess it maps ``Richard Erskine Leakey'' to
%% ``R. E. Leakey''; if instead the first input mapped to ``Dr. Jane'' or
%% ``Goodall, Jane'', we might guess the latter should map to
%% ``Dr. Richard'' or ``Leakey, Richard'', respectively.

%% The FlashFill system~\cite{gulwani2011automating} embedded in
%% Microsoft Excel solves problems such as these and is probably the best
%% known practical program-induction algorithm, but researchers in
%% programming languages and AI have had successes in many domains, such as handwriting recognition and
%% generation~\cite{lake2015human}, procedural
%% graphics~\cite{ellis2017learning,DBLP:journals/corr/abs-1804-01118}, question
%% answering~\cite{johnson2017clevr} and robot motion
%% planning~\cite{devlin2017neural}.  These systems
%% work in different ways, but most hinge upon a carefully engineered
%% \textbf{Domain Specific Language (DSL)}.  This is especially true for
%% systems like FlashFill that induce a wide range of programs
%% very quickly, in a few seconds or less.  DSLs constrain the search
%% over programs with strong prior knowledge in the form of a restricted
%% inventory of programming primitives finely tuned to the domain: for
%% text editing, these are operations like appending and
%% splitting on characters.

%% In this work, we consider the problem of building agents that learn to solve
%% program induction tasks, and also the problem of acquiring the prior knowledge
%% necessary to quickly solve these tasks in a new domain.  Representative problems
%% in three domains are shown in Table~\ref{initialExampleDSL}.  Our solution is an
%% algorithm that grows or boostraps a DSL while jointly training a neural network
%% to help write programs in the increasingly rich DSL.

\begin{table*}%[t!]
  \makebox[\textwidth][c]{ \scriptsize \tabcolsep=4pt
    \renewcommand\code\texttt \renewcommand\codechar[1]{\texttt{"#1"}}
    \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-0em}}c<{\hspace{-1em}}>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    &{\normalsize List Processing}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}&{\tabcolsep=7pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 2\, 3]}$\to $\code{[7\, 3]}         \\
          \code{[1\, 2\, 3\, 4]}$\to $\code{[3\, 4]} \\
          \code{[4\, 3\, 2\, 1]}$\to $\code{[4\, 3]} \\
          \pop{\code{$f(\ell) = $}\code{($f_1$ $\ell$ ($\lambda$ (x)}}\\
          \hspace{1.15cm}\pop{\code{(> x 2)))}}       \\
          \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \pop{\code{$f(\ell) = $}\code{($f_2$ $\ell$)}}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \pop{\code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{-}
      }}\\
      \hspace{1.25cm}\pop{\code{($f_0$ \codechar{.} \codechar{ }}}\\
      \hspace{1.4cm}\pop{\code{(cdr s)))}}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \pop{\code{$f($x$) = $($f_1$ x)}}&    \pop{\code{$f($x$) = $($f_6$ x)}}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \pop{\code{$f($x$) = $($f_4$ x)}}&    \pop{\code{$f($x$) = $($f_3$ x)}}\\

>>>>

\tikzstyle{vecArrow} = [thick, decoration={markings,mark=at position                                
   1 with {\arrow[semithick]{open triangle 60}}},                                                   
   double distance=1.4pt, shorten >= 5.5pt,                                                         
   preaction = {decorate},                                                                          
   postaction = {draw,line width=1.4pt, white,shorten >= 4.5pt}]                                    
\tikzstyle{innerWhite} = [semithick, white,line width=1.4pt, shorten >= 4.5pt]

@@

\begin{figure}
  \includegraphics[width = \textwidth]{figures/DeepLogo.pdf} 
  \end{figure}

An age-old dream within AI is a machine that learns and reasons by
writing its own programs.  This vision stretches back to the
1960's~\cite{solomonoff1964formal} and and is central to much of
what it would take to build machines that learn and think like humans.
Computational models of cognition often explain the flexibility and
richness of human thinking in terms of program learning: from everyday
thinking and problem solving (motor program induction as an account of
recognition and generation of handwriting and
speech~\cite{lake2015human}; functional programs as a model of natural
language semantics~\cite{SOMETHING}) to learning problems that unfold
over longer developmental time scales: the child's acquisition of
intuitive theories (of kinship, taxonomy, etc.)~\cite{Ullman2012} and
natural language grammar~\cite{DBLP:journals/cogsr/SchmidK11}, to name
just a few.  An outstanding challenge, however, is to engineer
program-learners that display the same level of domain-generality as
the humans they are meant to model.

We contribute a program-induction system that learns the
domain-specific concepts (DSL) while jointly learning how to use those
concepts.  This joint learning problem models two complementary
notions of domain expertise: (1) domain experts have at their disposal
a powerful, yet specialized repertoire of concepts and abstractions
(analogous to the DSL) while also (2) having accurate intuitions about
when and how to use those concepts to solve new problems.
Representative domains,
along with DSLs we learn for them,
are shown in Figure~\ref{exampleDSL}.

We call our system `DreamCoder' because it acquires these two kinds of
expertise through a novel kind of wake/sleep or `dream'
learning~\cite{hinton1995wake}, iterating through a wake cycle --
where it solves problems by writing programs -- and a pair of sleep
cycles, both of which are loosely biologically inspired by actual
human sleep.  The first sleep cycle, which we refer to as
\textbf{consolidation}, grows the DSL by replying experiences from
waking and consolidating them into new code abstractions.  This cycle
is inspired by the formation of abstractions during sleep memory
consolidation~\cite{DUDAI201520}.  The second sleep cycle, which we
refer to as \textbf{dreaming}, improves the agents knowledge of how to
write programs by training a neural network to help search for
programs. The neural net is trained on replayed experiences as well as
`fantasies', or samples, from the DSL.  These two kinds of dreams are
inspired by the distinct episodic replay and hallucination components
of dream sleep~\cite{fosse2003dreaming}.

This dream-learning architecture brings together two lines of prior work,
both of which have been separately influential within
artificial intelligence.
One line of work
considers the problem of learning new concepts, abstractions, or
`options' from
experience~\cite{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,solomonoff1989system,DBLP:journals/corr/abs-1110-5667,stolle2002learning},
while the other line of work considers the problem of learning how to
deploy those concepts
efficiently~\cite{devlin2017robustfill,balog2016deepcoder,NGDS}.  Our
goal with DreamCoder is to show that the combination of these ideas is
uniquely powerful, and pushes us toward program-writing systems that,
like human learners, autonomously acquire the expertise needed to
navigate a new domain of problems.

Because any learning problem can in principle be cast as program
induction, it is important to delimit our focus.  In contrast to
computer assisted programming~\cite{solar2008program} or genetic
programming~\cite{DBLP:books/daglib/0070933}, our goal is not to
automate software engineering, or to synthesize large bodies of code
starting from scratch.  Ours is a basic AI goal: capturing the human
ability to learn to think flexibly and efficiently in new domains ---
to learn what you need to know about a domain so you don't have to
solve new problems starting from scratch.  We are focused on problems
that people solve relatively quickly, once they acquire the relevant
domain expertise.  These correspond to tasks solved by short programs
--- if you have an expressive DSL.

    \newcommand{\name}[2]{
      \mcode{{\color{\inventionColor}$f_{#1}$}$($#2$)\,=\,$}
    }
    \newcommand{\noArguments}[1]{
      \mcode{{\color{\inventionColor}$f_{#1}$}$\,=\,$}
    }
    \newcommand{\use}[1]{{\color{\inventionColor}$f_{#1}$}}
    \renewcommand{\horizontalSpacing}{4.2cm}
    \renewcommand{\verticalSpacing}{2cm}

    \begin{scope}[shift={(0,8)}]
      \task{1}{2,3}{
        \textsc{Reverse}\\
        \mcode{[9 2 4]$\to$[4 2 9]}\\
        \mcode{[4 3]$\to$[3 4]}\\
        \mcode{[3 4 5 6]$\to$[6 5 4 3]}\\
        \mcode{[]$\to$[]}
      }{\footnotesize $f(\text{\mcode{z}}) = $\mcode{(fold ($\lambda$ (x a)}\\\hspace{0.25cm}\mcode{ (${\color{\inventionColor}f_3}$ x a)) nil z)}}
      \nextTask{2}{t1}{
        \textsc{Take if $\geq 5$}\\
        \mcode{[1 4 5 8]$\to$[5 8]}\\
        \mcode{[9 8 3 8]$\to$[9 8 8]}\\
        \mcode{[3 4 5 6]$\to$[5 6]}\\
        \mcode{[6 5 4 3]$\to$[6 5]}
      }{\footnotesize $f(\text{\mcode{z}}) = $\mcode{(${\color{\inventionColor}f_0}$ ($\lambda$ (x)}\\\hspace{0.5cm}\mcode{ (> x ${\color{\inventionColor}f_4}$)) z)}}

      %% \vinvention{filter}{p1}{
      %%   \name{0}{p$,$z}\mcode{\\(fold ($\lambda$ (x a)\\ (if (p x) (cons x a) a))\\ nil z)} 
      %% }{higher-order `filter'}
      \vinvention{single}{p1}{\name{3}{z$,$l}\\\mcode{(fold ($\lambda$ (x y)\\(cons x y))\\ z l)}}{Append element}

      \vinvention{three}{single}{\footnotesize \noArguments{1}\\\mcode{(+ 1 (+ 1 1))}}{The number `3'}
      \vinvention{prefix}{three}{\name{2}{n$,$z}\\\mcode{(map ($\lambda$ (i) (index i z)) (range (+ 1 n)))}}{First $n$ elements}
      \hinvention{filter}{single}{
        \name{0}{p$,$z}\mcode{\\(fold ($\lambda$ (x a)\\ (if (p x) (cons x a) a))\\ nil z)} 
      }{higher-order `filter'}
      \vinvention{for}{filter}{\noArguments{4}\\\mcode{(+ 1 ${\color{\inventionColor}f_1}$)}}{The number `4'}
      \vinvention{weird}{for}{\name{8}{p$,$l}\mcode{
          (map ($\lambda$ (x) (if (p x) (+ 1 x) 0)))}}{Higher-order incrementer}                 
      \hinvention{largest}{filter}{\name{6}{l$,$n}\\\mcode{(${\color{\inventionColor}f_5}$ (${\color{\inventionColor}f_0}$ ($\lambda$ (x) \\(> n (length (${\color{\inventionColor}f_0}$ ($\lambda$ (y) (> x y)) l)))) l))}}{$n^{\text{th}}$ largest element}
      \vinvention{Max}{largest}{\name{5}{l}\\\mcode{%
          (car (${\color{\inventionColor}f_0}$ ($\lambda$ (x) (nil? (${\color{\inventionColor}f_0}$ l ($\lambda$ (z) (> z x))))) l))}}{max of list}
      \vinvention{count}{Max}{\name{7}{p$,$l}\\\mcode{(length\\ (${\color{\inventionColor}f_0}$ ($\lambda$ (y) (= x y)) l))}}{count}

      \node[rotate=90,align=center] at ([xshift=-2cm]t1.center) {\textbf{\textsc{Tasks}}\\};
      \node(programLabel)[rotate=90,align=center] at ([xshift=-2cm]p1.center) {\texttt{\textbf{Progs.}}\\};
      \node[rotate=90,align=center] at ([xshift=-2cm]three.center) {\textbf{\textsc{Domain Specific Language}}\\
        \small Learned routines in {\color{orange}orange}};
      \node at ([yshift=0.5cm]t2.north) {\textbf{\Large \textsc{Domain: List Processing}}};

      \draw [dashed] ([xshift=0.5cm,yshift=-0.25\verticalSpacing]programLabel.east) -- ([xshift=15cm,yshift=-0.25\verticalSpacing]programLabel.east);
    \node[align=left,anchor=center] at ([xshift=14cm,yshift=-0.3\verticalSpacing]programLabel.east) {\large $\uparrow$\;system inputs\\\\\\\large $\downarrow$\;system outputs};

      \draw[->] (for.west) -- (three.east);
      \draw[->] (largest.east) to[out=0,in=0] (Max.east);
      \draw[->] ([yshift=-0.cm]largest.west) -- ([yshift=0.2cm]filter.east);
      \draw[->] (Max.west) -- (filter.east);
      \draw[->] ([yshift=0.1cm]count.west) -- ([yshift=-0.3cm]filter.east);

      \draw[->] ([yshift=0.35cm,xshift=-0.3cm]p2.south) to[out=250] (filter.north);
      \draw[->] ([yshift=0.0cm,xshift=0.3cm]p2.south) to[out=350,in=0] (for.east);
      \draw[->] ([xshift=-0.7cm]p1.south) -- (single.north);
      \draw[->] ([xshift=-1.3cm,yshift=0.6cm]p3.south) to[out=240] (largest.north);

    \node[align=center,rotate=90] at ([xshift=-2cm]t1.center) {\textbf{\textsc{Tasks}}\\};
    \node(programLabel)[align=center,rotate=90] at ([xshift=-2cm]p1.center) {\texttt{\textbf{Progs.}}\\};
    \node[rotate=90,align=center] at ([xshift=-2cm]pu.center) {\textbf{\textsc{Domain Specific Language}}\\
      \small Learned routines in {\color{orange}orange}};
          \node at ([yshift=0.5cm]t2.north) {\textbf{\Large \textsc{Domain: Graphics Programming}}};

    \draw [dashed] ([xshift=0.5cm,yshift=-0.25\verticalSpacing]programLabel.east) -- ([xshift=15cm,yshift=-0.25\verticalSpacing]programLabel.east);
    \node[align=left,anchor=center] at ([xshift=14cm,yshift=-0.3\verticalSpacing]programLabel.east) {\large $\uparrow$\;system inputs\\\\\\\large $\downarrow$\;system outputs};
  \end{tikzpicture}
  \caption{Two of the six domains we apply our system to. Agent observes tasks (top rows) which it solves by writing programs (middle rows) while jointly growing a library (DSL; bottom rows). Learned DSLs rediscover multiple higher-order functions (\code{filter} for list functions and rotational symmetry for generative graphics). Learned DSL components call each other (arrows).}  \label{exampleDSL}
\end{figure}

\section{Wake/Sleep Program Induction}\label{overviewSection}

\system~takes as its goal to acquire domain-specific expertise, which
it learns by solving a collection of programming \textbf{tasks}.  It
alternatingly finds programs that solve tasks (Wake --
Figure~\ref{threeCycles} top); improves its DSL by discovering and
reusing domain-specific subroutines (Consolidation --
Figure~\ref{threeCycles} left); and trains a neural network that
efficiently guides search for programs in the DSL (Dreaming --
Figure~\ref{threeCycles} right).  The learned DSL acts as a a prior on
programs likely to solve tasks in the domain, while the neural net
looks at a specific task and produces a ``posterior'' for programs
likely to solve that specific task (Figure~\ref{threeCycles} middle).
The neural network thus functions as a \textbf{recognition model}
supporting a form of approximate Bayesian program induction, jointly
trained with a \textbf{generative model} for programs encoded in the
DSL, in the spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL (and hence the search space for programs)
expands.  The generative model, or DSL, distills out common
abstractions across programs discovered during waking, growing a
network of increasingly deep and specialized domain-specific concepts
(Figure~\ref{exampleDSL}, bottom rows).

Viewed as an inference problem, our goal is to start with a collection
of tasks, written $X$, and infer both a program for each task, as well
as a distribution over programs encoded by a DSL, written
$\mathcal{D}$.  We equip $\mathcal{D}$ with a learned weight vector
$\theta$, and together $(\mathcal{D},\theta)$ define a generative
model over programs (see Appendix~\ref{sampleProgram}).  We perform
maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given
$X$.  Writing $J$ for the joint probability of $(\mathcal{D},\theta)$
and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]&\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta &\qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
where $\probability[x|p]$ scores the likelihood of a task
$x\in X$ given a program $p$.\footnote{For example, for list
  processing, the likelihood is 1 if the program predicts the observed
  outputs on the observed inputs, and 0 otherwise; when learning a
  generative model or probabilistic program, the likelihood is the
  probability of the program sampling the observation.}
The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $J(\mathcal{D},\theta)$ involves
summing over the  infinite set of all programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:

\begin{definition}
A \textbf{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs s.t. $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.
\end{definition}
\noindent Using the frontiers we  define the following intuitive lower bound on the joint probability, called $\lowerBound$:
\begin{align}
 J\geq \lowerBound\triangleq\probability[\mathcal{D},\theta]\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}
Our wake and sleep cycles correspond to alternate maximization of $\lowerBound$ w.r.t. $\left\{\mathcal{F}_x \right\}_{x\in X}$  (\textbf{Wake})
and $(\mathcal{D},\theta)$ (\textbf{Consolidation}):
\\\noindent \textbf{Wake: Maxing $\lowerBound$ w.r.t.\ the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x|p]\probability[p|\mathcal{D},\theta]\propto\probability[p|x,\mathcal{D},\theta]$ 
is large (i.e., programs with high posterior probability).
%Here we use a simple and generic enumerative program synthesis algorithm.
%% which we can accomplish by adding new programs to the frontiers means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{Sleep (Consolidation): Maxing $ \lowerBound$ w.r.t.\ the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed, and so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound \;\mathrm{d}\theta$,
and then updating $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $Q(\cdot |\cdot )$,
during the \textbf{Dreaming} phase: $Q$ is trained to approximate the
posterior over programs, $Q(p|x)\approx \probability[p|x,\mathcal{D}]\propto\probability[x|p]\probability[p|\mathcal{D}]$.
  Thus training the neural network amortizes the cost of finding programs with high posterior probability.

\noindent\textbf{Sleep (Dreaming): tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train %% a neural network, $q$, to predict a
%% distribution over programs conditioned on a task. The objective of $q$
%% is
$Q(p|x)$ to assign high probability to programs $p$ where
$\probability[x|p]\probability[p|\mathcal{D},\theta]$ is large, because incorporating those programs
into the frontiers will most increase $\lowerBound$.

Intuitively, this 3-phase inference procedure can work in practice because each of the 3 phases bootstraps off of the others. As the DSL grows and as the recognition model becomes more accurate, waking becomes more effective, allowing the agent to solve more tasks; when we solve more tasks during waking,
the Consolidation cycle has more data from which to learn the DSL; and, because the recognition model is trained on both
samples from the DSL and programs found during waking,
the recognition model gets both more data, and higher-quality data, whenever the DSL improves and whenever we discover more successful programs.
\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=white!5!white] (-1,1) -- (13,1) -- (13,-4) -- (-1,-4) -- (-1,1);
    \node at (5.5,1.25) {\textsc{\textbf{Wake}}};

@@

  \end{tikzpicture}
  \caption{\textbf{Middle:} \system as a graphical model. Agent observes programming tasks (e.g., input/outputs for list processing or images for graphics programs), which it explains with latent programs, while jointly inferring a latent Domain Specific Language (DSL) capturing cross-program regularities. A neural network, called the \emph{recognition model} (red arrows) is trained to quickly infer programs with high posterior probability. \textbf{Top}: Wake phase infers programs while holding the DSL and recognition model fixed. \textbf{Left}: Sleep (Consolidation) phase infers DSL while holding the programs fixed by refactoring programs found during waking and extracting common components. \textbf{Right}: Sleep (Dreaming) phase trains recognition model to predict approximate posterior over programs conditioned on task. Trained on `Fantasies' (programs sampled from DSL) \& `Replays' (programs found during waking).}\label{threeCycles}
\end{figure}

\subsection{Wake: Solving tasks}\label{explorationSection}

During waking we enumerate programs from the DSL in decreasing order
of their probability according to the recognition model, and then
check if a program $p$ assigns positive probability to a task
($\probability[x|p] > 0$); if so, we incorporate $p$ into the frontier
$\mathcal{F}_x$.  We represent programs as polymorphicly-typed
$\lambda$-calculus expressions, a representation closely resembling
Lisp and functional languages like Haskell and OCaml, including
variables, conditionals, higher-order recursive functions, and the
ability to create new functions.

%% Our programs are all
%% strongly typed.  We use the Hindley-Milner polymorphic typing
%% system~\cite{pierce} which is used in functional programming languages
%% like OCaml and Haskell.  We now define DSLs:
%% \begin{definition}
%% A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
%% A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
%% one number for each DSL element $e\in \mathcal{D}$, written $\theta_e$ and controlling the probability of  $e$ occurring in a program,
%% and a weight controlling the probability of a variable occurring in a program, $\theta_{\text{var}}$.
%% \end{definition}
%% \noindent Together with its weight vector, a DSL defines a distribution over
%% programs, $\probability[p|\mathcal{D},\theta]$.  We define this
%% distribution by specifying a procedure for drawing samples from
%% $\probability[p|\mathcal{D},\theta]$ (Appendix~\ref{generativeAppendix}).
%% Care must be taken to ensure that variable scoping rules are obeyed
%% and that programs are well-typed.  We ensure well-typed programs by
%% performing Hindley-Milner type inference~\cite{pierce} during
%% sampling, and assume that each task is annotated with the type of the
%% program that will solve it.
%% Appendix~\ref{enumerationAppendix}
%% explains how we enumerate,
%% rather than sample,
%% programs generated by Algorithm~\ref{sampleProgram}.
